---
title: "Organizing Data in R"
author: "Douglas Bates"
date: "2014-09-12"
output: ioslides_presentation
---

## Data frames
- Standard rectangular data sets (columns are variables, rows are observations) are stored in `R` as _data frames_.
- The columns can be _numeric_ variables (e.g. measurements or counts) or _factor_ variables (categorical data) or _ordered_ factor variables.  These types are called the _class_ of the variable.
- Many `R` packages contain sample data sets used to illustrate the techniques implemented in the package.
- There is also a `datasets` package containing datasets used in the example sections of the base `R` documentation.
    * many of these datasets are old and small, dating from the early days of `R`
```{r datasets}
ls("package:datasets")
```

## The `str` and `summary` functions
- The `str` function provides a concise description of the structure of a `data.frame` (or any other class of object in `R`).  The `summary` function summarizes each variable according to its class.  Both are highly recommended for routine use.
```{r strFormaldehyde}
str(Formaldehyde)
summary(Formaldehyde)
```

## `head` and `tail`
- Entering just the name of the data frame causes it to be printed.  For large data frames use the `head` and `tail` functions to view the first few or last few rows.
```{r headswiss}
head(OrchardSprays)
str(OrchardSprays)
```

## `ls.str`

- The operations of listing the objects in a package and providing a brief description of their structure are combined in `ls.str`
```{r lsstr}
ls.str("package:datasets")
```

## Data input
- The simplest way to input a rectangular data set is to save it as a comma-separated value (`csv`) file and read it with `read.csv`.
- The first argument is the name of the file.  On Windows it can be tricky to get the file path correct.  The `file.choose` function will bring up a chooser panel.
- `read.csv` just calls `read.table` with a different set of default arguments
- The first argument to `read.csv`, `read.table`, etc. can be a __connection__ or a __URL__ instead of a file name.
- Connection types (see `?connection`)
    * `gzfile` - a file compressed with `gzip`
    * `bzfile` - a file compressed with `bzip2`
    * `xzfile` - a file compressed with `xz`
    * `unz` - a single file from a zip archive

## Reading a compressed file or URL
```{r sd1,warning=FALSE}
str(sd1 <- read.csv(gzfile("./sd1.csv.gz","r")))
```
```{r classroom}
str(classroom <- read.csv("http://www-personal.umich.edu/~bwest/classroom.csv"))
```

## Copying, saving and restoring data objects
- Assigning a data object to a new name creates a copy.
- You can save a data object to a file, typically with the extension `.rda`, using the `save` function.
- To restore the data you `load` the file 
```{r saveload}
sprays <- InsectSprays
save(sprays,file="sprays.rda")
rm(sprays)
ls()
load("sprays.rda")
names(sprays)
```

