---
title: "Data visualization - `ggplot2`"
author: "Douglas Bates"
date: "2014-09-19"
output: ioslides_presentation
---
```{r preliminaries,cache=FALSE,echo=FALSE,results='hide'}
library(ggplot2)
library(knitr)
library(faraway)
opts_chunk$set(cache=TRUE,fig.align='center',fig.width=9,fig.height=4.5,warning=FALSE,message=FALSE)
```

# ggplot2

## The `ggplot2` graphics package
- `R` has "base" graphics functions as well as specialized plotting packages.
- The "base" graphics are a very old design.  Avoid them.  It is easy to create bad plots with them and difficult to create good graphics.
- The `ggplot2` package by Hadley Wickham of RStudio is the recommended visualization package.
- Hadley's book, [http://ggplot2.org](ggplot2: Elegant Graphics for Data Analysis) (Springer, 2009) describes the system.
- I will use data from the `faraway` package to accompany Julian Faraway's freely available book __Practical Regression and Anova using R__ to illustrate the use of `qplot`.

## General principles

- `ggplot2` is an implementation of Leland Wilkinson's "Grammar of Graphics" ideas
- The general approach is to build information about a plot in layers associating geometric or statistical aspects of the plot with various data characteristics.
- Frequently we create the base information about the plot then add information about the geometry then display the result.
- The `qplot` function produces a "quick plot".  It can be used to create a base plot but I prefer to use `ggplot`.

# The `pima` data set

## Examining the `pima` data

```{r pima}
str(pima)
head(pima)
```   

## Recoding the missing data
  
- As Faraway notes, several of the values of variables that cannot reasonably be zero are recorded as zero.

- A bit of research shows that these are missing data values. Also the `test` variable is a factor, not numeric.

```{r pimarecode}
pima <- within(pima, {
    diastolic[diastolic == 0] <- glucose[glucose == 0] <-
    triceps[triceps == 0] <- insulin[insulin == 0] <- bmi[bmi == 0] <- NA
    test <- factor(test, labels=c("negative","positive"))
})
head(pima, 3)
```

# Univariate summary plots

## Histogram of diastolic blood pressure
```{r histdiastolic}
qplot(diastolic, data=pima, geom="histogram")
```

## Comments about the histogram

- Histograms are from an earlier era.  You almost never want to use them.
- If you want to show the empirical density of a set of observations use an empirical density plot.
- One of the simplest things you can do to enhance a plot is to use informative axis labels, including the units if appropriate. In the case of diastolic blood pressure, "75 mmHg" is a measurement; "75" is not.
- To avoid repeating phrases like `xlab="Diastolic blood pressue (mmHg)"` create the base plot then display it with various geometries.

## Histogram of diastolic blood pressure
```{r histby}
p <- ggplot(pima, aes(x=diastolic)) + xlab("Diastolic blood pressure (mmHg)")
p + geom_histogram()
```   

## Empirical density plot of diastolic blood pressure
```{r density}
p + geom_density(aes(x=diastolic))
```   

## Empirical density of diastolic b.p. by test result
```{r densityby}
p + geom_density(aes(color=test))
```   

# Bivariate plots

## Simple scatterplot, c.f. Fig. 1.2a, p. 13
```{r scat}
p <- ggplot(pima,aes(x=diastolic,y=diabetes))+xlab("Diastolic blood pressure (mmHg)")+ylab("Diabetes pedigree function")
p + geom_point()
```   

## Adding a scatterplot smoother
```{r scat1}
p + geom_point() + geom_smooth()
```

## Multiple smoothers by group
```{r scat2}
p + geom_point(aes(color=test)) + geom_smooth(aes(color=test))
```   

## Conversion to log scale
```{r scat3}
p + geom_point(aes(color=test)) + geom_smooth(aes(color=test)) + scale_x_log10()
```

## Conversion to log-log scale
```{r scat4}
p <- p + geom_point(aes(color=test)) + geom_smooth(aes(color=test))
p + scale_x_log10() + scale_y_log10()
```

# Comparative boxplots

## Comparing a continuous response by groups
- One of the things that drive me crazy when I am looking at a paper or report is non-informative comparative boxplots.
- Almost always the response is shown on the vertical axis and the groups on the horizontal axis.
- By transposing these axes the plot can be made much more informative.
- In a report or paper, the "expense" of including a plot is usually determined by the height, not the width.
- Short wide plots are preferred to tall narrow plots.
- It makes much better sense to put the continuous response on the long axis and the categorical groups on the short axis.
- In `ggplot2` you create the plot in the usual orientation then use `coord_flip()` to transpose the axes.

## Comparative boxplots - the wrong way
```{r bw1}
p <- ggplot(pima, aes(x=test,y=diabetes)) + xlab("Diabetes test result") + ylab("Diabetes pedigree function")
p + geom_boxplot()
```   

## Comparative boxplot: horizontal orientation
```{r bw2,fig.height=1.5}
p + geom_boxplot() + coord_flip()
```
- You can make such a figure very short and not lose information

## Comparative boxplots: horizontal orientation, log scale
```{r bw3,fig.height=1.5}
p + geom_boxplot() + scale_y_log10() + coord_flip()
```   

# Simple regression or ancova lines

## Adding a simple linear regression line - c.f. Fig. 1.3
```{r scatlm,fig.height=4}
stat500 <- data.frame(scale(stat500))
p <- ggplot(stat500,aes(x=midterm,y=final))+geom_point()+xlab("Z-score on midterm exam")+ylab("Z-score on final exam")
p + geom_smooth(method="lm") + coord_equal()
```  

## Adding a reference line - c.f. Fig. 1.3
```{r scatlm1}
p+geom_smooth(method="lm")+geom_abline(intercept=0,slope=1,color="red")+coord_equal()
```

## Suppressing the confidence band
```{r scatlm2,fig.width=4.5}
p+geom_smooth(method="lm",se=FALSE)+geom_abline(color="red")+coord_equal()
```  

# Ancova

## Plotting multiple groups and lines, c.f. Fig. 15.2
```{r ancova}
levels(cathedral$style) <- c("Gothic", "Romanesque")
p <- ggplot(cathedral,aes(x=x,y=y)) + xlab("Nave Height (ft)") + ylab("Total Length (ft)")
p + geom_point(aes(color=style)) + geom_smooth(aes(color=style),method="lm")
```   

## Plotting multiple groups in separate panels
```{r ancova1}
p + geom_point() + geom_smooth(method="lm") + facet_grid(.~style)
```   
