---
title: "Introduction to the Hadleyverse"
author: "Douglas Bates"
date: "September 24, 2015"
output: 
  ioslides_presentation:
    fig_caption: yes
    fig_retina: null
    keep_md: yes
    smaller: yes
    widescreen: yes
---
```{r preliminaries,echo=FALSE,results='hide',cache=FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
options(width=102)
```

## RStudio and Hadley Wickham

- The biggest changes in the __R__ environment over the last 10 years have come from [Rstudio](http://rstudio.com)
  and its chief scientist [Hadley Wickham](https://github.com/hadley)

- Hadley did his Ph.D. at Iowa State, during which he developed the [ggplot2](https://github.com/hadley/ggplot2) package implementing the "Grammar of Graphics"

- He continues to be a driving force in innovation for the __R__ environment, to the extent that his collection of packages is known as the _Hadleyverse_

- He was also an early adopter of [github](https://github.com) and through its integration with RStudio encouraged its use with __R__

- The packages he has written are characterized by composability
    * [readr](http://github.com/hadley/readr) - replacements for data-reading functions suitable for use with 
    * [dplyr](https://github.com/hadley/dplyr) - data manipulation, grouping and summaries
    * [tidyr](https://github.com/hadley/tidyr) - transfer from `wide` to `long` format and vice-versa
    * [rvest](https://github.com/hadley/rvest) - web-scraping in R
    * [devtools](https://github.com/hadley/devtools) - utilities for building and maintaining __R__ packages

## [readr](http://github.com/hadley/readr) and [dplyr](https://github.com/hadley/dplyr)

Earlier we read the `classroom` data using
```{r classroom,eval=FALSE}
str(classroom <- read.csv("http://www-personal.umich.edu/~bwest/classroom.csv"))
```
The equivalent expression using  [readr](http://github.com/hadley/readr) is
```{r classroomreadr}
(classroom <- read_csv("http://www-personal.umich.edu/~bwest/classroom.csv"))
```
Note that `readr` function calls produce `dplyr` tables.



