---
title: "Simulating linear model fits"
author: "Douglas Bates"
date: "11/12/2014"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: lualatex
    number_sections: yes
    toc: yes
---
```{r preliminaries,echo=FALSE,results='hide'}
library(knitr)
library(ggplot2)
opts_chunk$set(fig.align='center',fig.pos="tb",cache=TRUE)
```

# Simulating linear model fits

In a model-fitting function, such as `lm` or `glm`, a large proportion
of the execution time can be spent on decoding and checking the
argument values, creating the model frame (function `model.frame`) and
the model matrix (function `model.matrix`) and preparing the arguments
for the numerical operations, which are performed in `lm.fit` or
`glm.fit`. 

This is fine when we are fitting a single model or performing model
building.  The time we should spend thinking about the model fits
dwarfs the time spent fitting the model, in most cases.

However, when we are going to perform a simulation, we don't want to
perform all these operations of setting up the model when we are
fitting the same model to tens of thousands of simulated response
vectors in a simulation study.

## Fitting a matrix of responses

Fortunately there is a very fast, but little known, way of fitting a
large number of simulated responses to the same model.  If you read
the documentation for the `lm` function you will find that the
response part of the formula can be a matrix with n rows and N
columns.  That is, you can fit all the simulated responses in a single
call to `lm`
```{r bigMod}
set.seed(1234321)            # ensure a reproducible stream
N <- 10000                   # number of replications
n <- 22                      # size of the sample
dat <- data.frame(x=1:n)     # x values for a simple linear regression
(betaTrue <- c(0, 4) + rnorm(2, sd = 0.1))

muTrue <- as.vector(model.matrix(~ x, dat) %*% betaTrue)
str(Ymat <- matrix(rnorm(n*N, mean=0, sd=0.3), nr = n) + muTrue)
system.time(bigMod <- lm(Ymat ~ x, dat))
```

As you can see, this is very fast -- less than 1 second which is much,
much faster than any kind of looping or apply function could achieve.

The usual extractors applied to such model fits produce vectors or
matrices.
```{r coefbigmod}
str(cc <- coef(bigMod))
```

To plot the distribution of the coefficient estimates we create a data
frame of the transpose of this matrix.
```{r}
coefFr <- as.data.frame(t(cc))
names(coefFr) <- c("Intercept", "slope")
```
```{r beta0dens,echo=FALSE,fig.cap="Empirical density plot of intercept coefficient from 10,000 simulations of a simple linear regression"}
qplot(Intercept, data=coefFr, geom="density", xlab=expression(hat(beta)[0]))
```
```{r beta1dens,echo=FALSE,fig.cap="Empirical density plot of intercept coefficient from 10,000 simulations of a simple linear regression"}
qplot(slope, data=coefFr, geom="density", xlab=expression(hat(beta)[1]))
```


In a scatter plot of a large number of points such as we have here, we sometimes
get a better impression of the density of the points by using "alpha
blending".  We set the parameter `alpha` to a value between 0 and 1 so
the points are partially transparent.  (There is a reason we give the value of
`alpha` as a quoted string.  Try without the quotes to see the difference.)

```{r alphaplot,fig.cap="Scatterplot of estimated coefficients from 10,000 simulations of a simple linear regression"}
qplot(Intercept, slope, data=coefFr, alpha="0.2",
      xlab=expression(hat(beta)[0]),ylab=expression(hat(beta)[1]))
```
